Discrete Ordinates

### 2017-05-12 docprep/chapbib.tex
Worked on the intro and results

 %% ... sample chapter ...

This chapter summarizes the literature reviewed in preparation of this work. The literature review is split into multiple sections, each highlighting relevant works pertaining to a particular component or section of this work. The first sections cover introductory work and steps necessary for preprocessing data. Section~\ref{sec:discordlit} covers historical and recent work in discrete ordinate methodology, the methodology used by DOCTORS. The remaining sections cover GPU hardware and other implementation facets.

@@

@@

\section{CT Phantom Generation}
The CT phantom is populated with CT numbers, also known as Hounsfield numbers. These values are related to the attenuation coefficient of the material represented by the voxel.

@@

The new guy provides 19 materials. The density is measured.

@@

\section{Discrete Ordinates}\label{sec:discordlit}
Discrete ordinates are old. The book is \citet{ref:lewise}.

>>>>

This chapter summarizes the literature reviewed in preparation of this work. The literature review is split into multiple sections, each highlighting relevant works pertaining to a particular component or section of this work. The first two sections cover introductory work and steps necessary for preprocessing data. Section~\ref{sec:discordlit} covers historical and recent work in discrete ordinate solution methodology, the methodology used by DOCTORS. The remaining sections cover GPU hardware and other implementation facets.

@@

MOTIVATION

@@

COMPUTATION

\section{CT Phantom Generation}
The CT phantom is populated with CT numbers, also known as Hounsfield numbers. These values are related to the attenuation coefficient of the material represented by the voxel. Traditionally, a CT number of zero corresponds to the attenuation of water while -1000 corresponds to dry air CITE though some variations do exist CITE.

@@

The new guy provides 19 materials. The density is measured.
The DOCTORS code relies on the CT nuber to material mapping of \citet{ref:ottossonr} who developed a 19-group BLAH. Table~\ref{table:ctmap} shows the mapping of CT number to material composition (reproduced from \citep{ref:ottossonr}).

\begin{table}[ht]
\caption{Water Regions}
\centering 
\begin{tabular}{l c r r r r r r r r r r r r}
\hline \hline   
Media       & CT Range      & H    & C    & N    & 0    & Na  & Mg & P    & S   & Cl  & Ar  & K   & Ca \\ [0.5ex] 
\hline
Air         & -950 to -100  &      &      & 75.7 & 23.3 &     &    &      &     &     & 1.3 &     &       \\
Lung        & -1000 to -950 & 10.3 & 10.5 & 3.1  & 74.9 & 0.2 &    & 0.2  & 0.3 & 0.3 &     & 0.2 &       \\
Adipose     & -100 to 15    & 11.2 & 50.8 & 1.2  & 36.4 & 0.1 &    &      & 0.1 & 0.1 &     &     &       \\
Connective  & 15 to 129     & 10.0 & 16.3 & 4.3  & 68.4 & 0.4 &    &      & 0.4 & 0.3 &     &     &       \\
Bone        & 129 to 200    & 9.7  & 44.7 & 2.5  & 35.9 &     &    & 2.3  & 0.2 & 0.1 &     & 1.0 & 4.5   \\  [1ex]
\hline
\end{tabular}
\label{table:ctmap}
\end{table}

@@

\section{Discrete Ordinates}\label{sec:discordlit}
The dsicrete ordinates method dates back to CITE but remains one of the most prominent solution modalities for ration transport in use today.

A comprehensive review of discrete ordinate methods is given by \citet{ref:lewise}. Additional references include

### 2017-05-12 docprep/chapintro.tex
Added a new chapter - results.
Worked on the intro and results

Another application is in treatment.

Currently, no algorithm exists to provide this information.

The goal of this work is to provide such an algorithm.

There are two primary avaialble for the methodologies: Monte Carlo and deterministic methods. Monte Carlo has proven to be very slow in these areas. Furhter, the 3D Cartesian mesh generated from CT reconstruction is well suited to deterministic computation.

@@

>>>>

Currently, dose estimation relies on \textit{a priori} computation verified with a standardized benchmark. No methodology currently exists to verify that the dosimetry evaluation was accurate after the patient has undergone the procedure. This work proposes a methodology by which a patient's CT reconstruction is used to compute the dose received from the radiation beam.

Another, similar, application is dose estimation of patients receiving radiation therapy. Often, before the procedure, a time dependent, 4D CT scan of the patient is taken so that radiologists can account for breathing patterns during administration of the treatment.

ESTIMATE. PHANTOM. CORRELATE.

BASED ON REFERENCE PHANTOM.

ADD FIGURE

Once the procedure (whether diagnostic or treatment) is completed, the dose distribution is assumed to follow the empirical model.

There are two primary methodologies available to provide the transport solution: Monte Carlo and deterministic methods. Monte Carlo has proven to be very slow in these areas. Further, the 3D Cartesian mesh generated from CT reconstruction is well suited to deterministic computation.

@@

Computational mesh from a sinogram. Reconstruction.
The DOCTORS code takes a computational reconstruction mesh of CT numbers, cross section data, and other parameters related to the solution methodology as input. The output is the collided and uncollided flux as well as the dose in each computational cell. These values can be compared to the empirical correlation results as a verification that the patient received an appropriate dose.

SHOULD BE FAST - GPU

USER INTERFACE

### 2017-05-12 docprep/chapmeth.tex
Worked on the intro and results

\section{Mathmatic Development}

Where does this text go?

@@

The time dependent gamma transport equation can be written as

\begin{equation} \label{eq:boltz_t_dep}
\begin{split}
	&\left[ \frac{1}{v(E)} \frac{\partial}{\partial t} + \hat{\Omega} \cdot \nabla + \Sigma_t(\boldsymbol{r}, E, t) \right]
	\psi(\boldsymbol{r}, E, \hat{\Omega}, t) = \\
	&\int_{4 \pi} \int_0^\infty \Sigma_s(\boldsymbol{r}, E' \rightarrow E, \hat{\Omega}' \rightarrow \hat{\Omega}, t) \psi(\boldsymbol{r}, E', \hat{\Omega}', t) dE' d\hat{\Omega}' + S(\boldsymbol{r}, E, \hat{\Omega}, t)
\end{split}
\end{equation}

However, we are typically not concerned with the transient case in medical diagnostic imaging. We are more interested in the steady state case. The time independent form of Eq. \ref{eq:boltz_t_dep} is written as

@@

The following sections show how Eq.~\ref{eq:boltz} is discretized.

\subsection{Angular Discretization}
The coordinate system used in DOCTORS is shown in Fig.~\ref{fig:coord_sys}. A discrete set of $N_a$ angles ($\Omega_{a} \quad a = 0 \ldots N_a-1$) is selected to represent continuous directional space. Particles are transported only along these discrete directions form one voxel to adjacent voxels.

@@

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/coord_sys}
  \end{center}
  \caption{The coordinate system used in DOCTORS. Given an arbitrary direction, $\Omega$, $\mu$, $\eta$, and $\xi$ are its direction cosines with respect to the $x$, $y$, and $z$ axes respectively. $\varphi$ is the azimuthal angle (with respect to $x$ and the d $\theta$ is the polar angle (with respect to $z$).}
\label{fig:coord_sys}
\end{figure}%

@@

@@

\section{Uncollided Flux}
To avoid ray effect, the flux is split into collided and uncollided components as shown in Eq.~\ref{eq:totflux}. Equation~\ref{eq:boltz} can be rewritten as two equations solving for the uncollided flux, $\psi_u$, and collided flux, $\psi_c$ independently. Equation~\ref{eq:unc} has no scatter term since scattered particles are not considered in the uncollided flux. The source term, $S_u$, in Eq.~\ref{eq:col} is the first collision source.

@@

>>>>

This chapter summarizes the solution methodology employed by DOCTORS. The techniques used to compute the collided and uncollided fluxes are described in seperate sections. After the flux methodology is explained, the methods used for source generation and covered as well as the flux to dose conversion.

\section{Collided Solution Methods}

The discrete ordinate solver computes the collided flux distribution inside the CT mesh given an input uncollided flux, physical geometry information, and other associated solver parameters such as quadrature and energy discretization.

@@

The time dependent gamma transport equation is

\begin{equation} \label{eq:boltz_t_dep}
\begin{split}
	&\left[ \frac{1}{v(\boldsymbol{r}), t} \frac{\partial}{\partial t} + \hat{\Omega} \cdot \nabla + \Sigma_t(\boldsymbol{r}, E, t) \right]
	\psi(\boldsymbol{r}, E, \hat{\Omega}, t) = \\
	&\int_{4 \pi} \int_0^\infty \Sigma_s(\boldsymbol{r}, E' \rightarrow E, \hat{\Omega}' \rightarrow \hat{\Omega}, t) \psi(\boldsymbol{r}, E', \hat{\Omega}', t) dE' d\hat{\Omega}' + S(\boldsymbol{r}, E, \hat{\Omega}, t)
\end{split}
\end{equation}

where $v$ is the velocity of the photon in the medium at $\boldsymbol{r}$, $\Sigma_t$ is the total macroscopic cross section, $\Sigma_s$ is the macroscopic scattering cross section, $S$ is an external photon source, and $\phi$ is the total photon angular flux. However, the transient case is not typically of interest in medical diagnostic imaging, rather the steady state is more important. The time dependence in Eq. \ref{eq:boltz_t_dep} is removed to yield

@@

The following sections show how Eq.~\ref{eq:boltz} is discretized and solved computationally.

\subsection{Angular Discretization}
The coordinate system used in DOCTORS is shown in Fig.~\ref{fig:coord_sys}. A discrete set of $N_a$ angles ($\Omega_{a}, \quad a = 0 \ldots N_a-1$) is selected to represent continuous directional space. Particles are transported only along these discrete directions from one voxel to adjacent voxels.

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/coord_sys}
  \end{center}
  \caption{The coordinate system used in DOCTORS. Given an arbitrary direction, $\Omega$, $\mu$, $\eta$, and $\xi$ are its direction cosines with respect to the $x$, $y$, and $z$ axes respectively. $\varphi$ is the azimuthal angle (with respect to $x$ and the d $\theta$ is the polar angle (with respect to $z$).}
\label{fig:coord_sys}
\end{figure}

@@

The plain text file should be formatted such that it contains four columns similar to Table~\ref{tab:quad_format} (without any headings). No checks are performed to ensure that octants are covered, this allows simple testing quadratures that utilize only a single direction to be used for testing purposes. Values for $\mu$, $\eta$, and $\xi$ do not necessarily need to add to unity, they will be automatically normalized by DOCTORS. Likewise, the weights will be normalized such that their summation is unity.

\begin{table}[ht]
\caption{User Defined Quadrature Format}
\centering 
\begin{tabular}{c c c c}
\hline \hline   
$\mu$    & $\eta$ & $\xi$ & Weight\\ [0.5ex] 
\hline
5.12345  & 5.12345 & 5.12345 & 5.12345 \\
5.12345  & 5.12345 & 5.12345 & 5.12345 \\
5.12345  & 5.12345 & 5.12345 & 5.12345 \\
5.12345  & 5.12345 & 5.12345 & 5.12345 \\
5.12345  & 5.12345 & 5.12345 & 5.12345 \\
5.12345  & 5.12345 & 5.12345 & 5.12345 \\
5.12345  & 5.12345 & 5.12345 & 5.12345 \\
5.12345  & 5.12345 & 5.12345 & 5.12345 \\ [1ex]
\hline
\end{tabular}
\label{table:quad_format}
\end{table}

@@

Once the data is parsed, weighted combinations of elemental data is used to create material cross sections.

WATER EXAMPLE.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/airwaterxs1_19group}
        \caption{Data}
        \label{fig:ampx1}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/airwaterxs2_19group}
        \caption{Same data.}
        \label{fig:ampx2}
    \end{subfigure}
    \caption{Comparison to NIST data.}\label{fig:airwaterxs_19group}
\end{figure}

@@

\section{Uncollided Solution Methods}
To avoid ray effect, the flux is split into collided and uncollided components as shown in Eq.~\ref{eq:totflux}. Equation~\ref{eq:boltz} can be rewritten as two equations solving for the uncollided flux, $\psi_u$, and collided flux, $\psi_c$ independently. Equation~\ref{eq:unc} has no scatter term since scattered particles are not considered in the uncollided flux. The source term, $S_u$, in Eq.~\ref{eq:col} is the first collision source.

@@

\section{Dose Computation}
Dose is computed.

### 2017-05-12 docprep/chapresults.tex
Worked on the intro and results

@@

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/waterHistogramLin2}
        \caption{Linear}
        \label{fig:waterHistLin}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/waterHistogramLog}
        \caption{Logarithmic}
        \label{fig:waterHistLog}
    \end{subfigure}
    \caption{The histogram.}\label{fig:waterHist}
\end{figure}

\subsection{Gantry Handling}
The gantry is there.

\section{Benchmark}

MCNP is good.

\section{Comparison}

They agree.

>>>>

First subtract 1024 to account for the unsigned 16-bit int data structure.

@@

\subsection{Gantry Handling}
In some phantom models, a gantry is present. The gantry poses a challenge since it is a highly attenuating material relative to the patient which reduces the available flux.

The CT number to material conversion developed by~\cite{ref:ottossonr} which is implemented in this work does not have an equivalent for the gantry region.

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/waterPhantom256}
  \end{center}
  \caption{Raw CT number data for the $256 \times 256 \times 64$ water phantom.}
\label{fig:waterPhantom256}
\end{figure}

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/waterPhantom64}
  \end{center}
  \caption{The small one}
\label{fig:waterPhantom64}
\end{figure}

\section{Cone-Beam Water Benchmark}

The first benchmark, the cone with 19 groups is given in Fig. 1. The cross section data for the water and air regions are shown in Fig.~\ref{fig:airwater19}.

The water phantom is a 35 cm diameter water cylinder that stands 12.5 inches tall. The water is encased in a 0.5 cm thick plastic container.

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/waterHistogramLin}
        \caption{Linear}
        \label{fig:waterHistLin}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/waterHistogramLog}
        \caption{Logarithmic}
        \label{fig:waterHistLog}
    \end{subfigure}
    \caption{The histogram.}\label{fig:waterHist}
\end{figure}

Due to the meshing used in CT phantoms, the spatial domain of the problem is identical for both the deterministic and Monte Carlo problems. A key difference between though is the group structure used in the group-averaged deterministic cross sections. Figure~\ref{fig:airwarerxs} shows the continuous energy macroscopic cross section for both air and water versus the discretized data used by DOCTORS. Unfortunately, the datasets used in DOCTORS originate from SCALE6.2, a light water reactor analysis code. Therefore, their upper photon energy bound is 20 MeV which is far higher than necessary for diagnostic imaging. This results in the region of interesting being discretized very roughly. In the 19-group discretization, the entire diagnostic domain is covered by only three energy groups.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/airwaterxs2_19group}
        \caption{19 Groups}
        \label{fig:airwaterxs19}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/airwaterxs2_47group}
        \caption{47 Groups}
        \label{fig:airwaterxs47}
    \end{subfigure}
    \caption{Comparison of the group cross section data used by DOCTORS to the continuous cross section data used by MCNP6. The continuous data is provided by NIST !!!CITE!!!.}\label{fig:airwaterxs}
\end{figure}

The detailed flux distribution from MCNP6 and the solver are not included here, they are provided in Appendix~\ref{sec:appc_benchmark}.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/bench_cone1_comp}
        \caption{10 - 20 MeV.}
        \label{fig:bench_cone1_comp}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/bench_cone1_lineout}
        \caption{10 - 20 MeV lineout.}
        \label{fig:bench_cone1_lineout}
    \end{subfigure}

    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/bench_cone1_comp_g17}
        \caption{100 - 200 keV.}
        \label{fig:bench_cone1_comp_g17}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/bench_cone1_lineout_g17}
        \caption{100 - 200 keV lineout.}
        \label{fig:bench_cone1_lineout_g17}
    \end{subfigure}

    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/bench_cone1_comp_g18}
        \caption{45 - 100 keV.}
        \label{fig:bench_cone1_comp_g18}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/bench_cone1_lineout_g18}
        \caption{45 - 100 keV lineout.}
        \label{fig:bench_cone1_lineout_g18}
    \end{subfigure}

    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/bench_cone1_comp_g19}
        \caption{10 - 45 keV.}
        \label{fig:bench_cone1_comp_g19}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/bench_cone1_lineout_g19}
        \caption{10 - 45 keV lineout.}
        \label{fig:bench_cone1_lineout_g19}
    \end{subfigure}
    \caption{Comparison of the MCNP6 results and the raytracer.}\label{fig:waterHist}
\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/bench_cone1_centerdiff}
        \caption{}
        \label{fig:bench_cone1_centerdiff}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/bench_cone1_centerdiff2}
        \caption{}
        \label{fig:bench_cone1_centerdiff2}
    \end{subfigure}
    \caption{Comparison}\label{fig:bench_cone1_centerdifffig}
\end{figure}

\section{3 Cone-Beam Water Benchmark}

Now there are three

\section{16 Cone-Beam Water Benchmark}

And now 16.

\section{1 Cone-beam Phantom Benchmark}

It's a person!

\section{16 Cone-beam Phantom Benchmark}

It's a person!

### 2017-05-12 docprep/norris_phd_dissertation.tex
Added more info on the subsweep
Added a new chapter - results.
Worked on the intro and results

 % -*- mode: LaTeX -*-
 % ... preamble ...
\documentclass[times,12pt,titlepage]{mstthesis}
\doublespacing
\usepackage{threeparttable}
\usepackage[final]{graphicx}
\usepackage{psfrag}
\usepackage{listings}
\usepackage{minted}
\usepackage[noprefix]{nomencl}
\makenomenclature
\renewcommand{\nomgroup}[1]{
   \ifthenelse{\equal{#1}{A}}{\medskip \item \textbf{Roman}}{}%
   \ifthenelse{\equal{#1}{G}}{\medskip \item \textbf{Greek}}{}%
   \ifthenelse{\equal{#1}{S}}{\medskip \item \textbf{Subscripts}}{}%
}
\usepackage[nopar]{lipsum} % ... provides dummy text ...

@@

\begin{ThesisAcknowledgment}
I would like to thank these people. They are all so super cool. So much coolness. So much so that this is now over a line long.

And these funding agencies.
\end{ThesisAcknowledgment}
\begin{ThesisFrontMatter}
\tableofcontents
\listoffigures
\listoftables
\listofsymbols
\end{ThesisFrontMatter}

@@

\ThesisBodyChapter{Implementation}\input{chapimpl}
 % ... conclusions chapter ...
\ThesisBodyChapter{Conclusions}\input{chapconc}

@@

\begin{ThesisAppendix}{two}
\ThesisAppendixChapter{Legendre Polynomials and Spherical Harmonics}\input{appdx_a}
\ThesisAppendixChapter{Two}\input{appdx_b}
\end{ThesisAppendix}

@@

\begin{Vita}
Edward Norris graduated from Missouri University of Science and Technology in 2013 with a B.S. in both nuclear engineering and computer science.

He has worked...

He is a member of...
\end{Vita}

>>>>

 % -*- mode: LaTeX -*-
 % ... preamble ...
\documentclass[times,12pt,titlepage]{mstthesis}
\doublespacing
\usepackage{threeparttable}
\usepackage[final]{graphicx}
\usepackage{psfrag}
\usepackage{listings}
\usepackage{minted}
\usepackage[noprefix]{nomencl}
\usepackage{subcaption}
\usepackage{xr}
\makenomenclature
\renewcommand{\nomgroup}[1]{
   \ifthenelse{\equal{#1}{A}}{\medskip \item \textbf{Roman}}{}%
   \ifthenelse{\equal{#1}{G}}{\medskip \item \textbf{Greek}}{}%
   \ifthenelse{\equal{#1}{S}}{\medskip \item \textbf{Subscripts}}{}%
}
\usepackage[nopar]{lipsum} % ... provides dummy text ...

@@

\begin{ThesisAcknowledgment}
I would like to thank my adviser, Dr. Liu and all of the other graduate students for their continued support through my graduate career. I also appreciate the time spent by my committee members helping me identify problems and provide insight to solutions.

I would also like to acknowledge the NRC for providing funding for my work. My work is supported by NRC grant NRC-HQ-13-G-38-0026.
\end{ThesisAcknowledgment}

\begin{ThesisFrontMatter}
\tableofcontents
\listoffigures
\listoftables
\listofsymbols
\end{ThesisFrontMatter}

@@

\ThesisBodyChapter{Implementation}\input{chapimpl}
 % ... conclusions chapter ...

\ThesisBodyChapter{Results}\input{chapresults}

\ThesisBodyChapter{Conclusions}\input{chapconc}

@@

\begin{ThesisAppendix}{two}
\ThesisAppendixChapter{Legendre Polynomials and Spherical Harmonics}\input{appdx_a}

\ThesisAppendixChapter{Code Listings}\input{appdx_b}

\ThesisAppendixChapter{Benchmark Results}\input{appdx_c}

\end{ThesisAppendix}

@@

\begin{Vita}
Edward Norris graduated from Missouri University of Science and Technology in 2013 with a B.S. in both nuclear engineering and computer science, he went on to pursue a M.S. in computer science and a PhD in nuclear engineering.

His PhD work was funded by a fellowship through the NRC.

Previously he has participated in four summer internships with Sandia National Laboratories. While there, he served in various capacities ranging from cyber defense to radiation detection and neutron scatter simulation.

Edward was awarded SOME AWARDS.

Edward was also an active member of the Nuclear Science Design Team at Missouri University of Science and Technology. While on the team he solved radiation transport problems to facilitate the dosimetry of an inertial electrostatic confintement fusion device being constructed.

\end{Vita}
